An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-07-11 nips_2016.tex
added more test scripts

In the Metropolis-Hastings method~\cite{gilks1996markov,brooks2011handbook}, a difficult-to-compute
probability distribution $p(\theta)$ is sampled using a Markov chain $\theta_1,\ldots,\theta_n$. The 
sample $\theta_{t+1}$ at time $t+1$ is generated using a candidate $\theta'$ from a (simpler) proposal
distribution $q(\theta'\mid \theta_t)$, filtered by an acceptance test. The acceptance test is usually
a Metropolis test. The Metropolis test has acceptance probability:
\begin{equation}\label{eq:traditional}
\alpha(\theta_t,\theta') = \min\left\{ 1, \frac{p(\theta')q(\theta_t \mid \theta')}{p(\theta_t)q(\theta' \mid \theta_t)}
\right\}
\end{equation}
and with probability $\alpha(\theta_t,\theta')$, we accept $\theta'$
and set $\theta_{t+1} = \theta'$, otherwise set
$\theta_{t+1}=\theta_t$.  The test is often implemented with an
auxiliary random variable $u \sim \mathcal{U}(0,1)$ with a comparison $u < \alpha(\theta_t,\theta')$.
For notational simplicity, from now on we will drop the subscript $t$ for the current sample $\theta_t$ and
denote is simply as $\theta$.

The acceptance test guarantees detailed balance, which means
\begin{equation}\label{detailed_balance1}
  p(\theta)p(\theta,\theta') = p(\theta')p(\theta',\theta)
\end{equation}
where $p(\theta,\theta')$ is the probability of a transition from state $\theta$ to $\theta'$. Here
$p(\theta,\theta') = q(\theta'\mid\theta)\alpha(\theta,\theta')$ so the detailed balance equation becomes
\begin{equation}\label{detailed_balance2}
  p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta)
\end{equation}
The detailed balance condition, together with ergodicity, guarantee that the
Markov chain has a unique stationary distribution $\pi(\theta) = p(\theta)$.

For Bayesian inference, we would like to sample from a parameter distribution for $\theta$ based on some
observed data $x_1,\ldots,x_N$. i.e. we want to sample from 
$p(\theta \mid x_1, \ldots, x_N)$. The acceptance probability is now:
\begin{equation}
\alpha(\theta,\theta') = 
\min\left\{ 1, \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
\theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)} \right\},
\end{equation}
where $p_0(\theta)$ is a prior, and $p(x_i \mid \theta)$ are the
probabilities of the observations. Computing samples this way requires
the use of all $N$ training data points, but this is very expensive
for large datasets. To address this challenge,~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate Metropolis-Hasting tests using sequential hypothesis
testing. During each iteration, they start with a small minibatch of
data and test the hypothesis that the sample $\theta'$ based on an
approximate version of the test $u < \alpha(\theta,\theta')$. If the
approximate test cannot make a decision with sufficient confidence,
then the minibatch size is increased and the test repeats. This
process continues until a decision. The bounds depend on either a
asymptotic central limit theorem \cite{cutting_mh_2014} or a
concentration bound \cite{icml2014c1_bardenet14}, the latter requiring
direct bounds on the log likelihood ratio.  The problem with both
methods is the expense of resolving small differences.  In the worst
case, all $N$ data points may be needed, and as we show next this
worst case can occur about $\Omega(N)$ times during the performance of
$N$ tests.

Following ~\cite{icml2014c1_bardenet14}, we write the test $u < \alpha(\theta,\theta')$ in
the equivalent form $\Lambda(\theta,\theta') > \psi(u,\theta,\theta')$ where
\begin{equation}
\Lambda(\theta,\theta') = \frac{1}{N}\sum_{i=1}^N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\hspace{0.2in} {\rm and} \hspace{0.2in}
\psi(u,\theta,\theta') = \frac{1}{N}\log\left(u\frac{q(\theta'|\theta)p_0(\theta)}{q(\theta|\theta')p_0(\theta')}\right)
\end{equation}

@@

\begin{equation}
\Lambda^*(\theta,\theta') = \frac{1}{b}\sum_{i=1}^b \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}

The minibatch should contain randomly selected samples, and
by the central limit theorem, we expect $\Lambda^*(\theta,\theta')$ to be approximately
Normal. 
The acceptance test then becomes a statistical test of the hypothesis
that $\Lambda(\theta,\theta') > \psi(u,\theta,\theta')$ by
establishing that $\Lambda^*(\theta,\theta')$ is substantially larger
than $\psi(u,\theta,\theta')$.  In ~\cite{cutting_mh_2014} an
asymptotic central limit argument was used to derive this gap, while
in \cite{icml2014c1_bardenet14} a concentration bound was used. In
both cases, the resulting tests were shown to give useful reductions
in number of samples required over using the full dataset. But no
worst-case bounds were given on average number of samples needed.

@@

\subsection{An Example}
Consider a simple model where samples $x_i$ for $i=1,\ldots,N$ are
drawn from a Normal distribution $\mathcal{N}(\theta,1)$ where
$\theta=0.5$, and with a uniform prior on $\theta$. The log likelihood ratio is
\begin{equation}
\Lambda^*(\theta,\theta') =  \frac{1}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta)}=
  %%\frac{1}{2}\left( (x_i-\theta')^2 - (x_i - \theta)^2\right) =
  (\theta'-\theta)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.
Since the $x_i$ have unit variance, their mean has variance $1/b$, and the variance $\sigma^2(\Lambda^*)$
of $\Lambda^*(\theta,\theta')$ is $(\theta'-\theta)^2/b$. In order to pass
a hypothesis test that $\Lambda > \psi$, we will need to establish a
large enough gap (several $\sigma(\Lambda^*)$) between $\Lambda^*(\theta',\theta)$
and $\psi(u,\theta',\theta)$.

We would like to sample from the posterior distribution of $\theta$
which is a normal distribution centered on the sample mean, and with
variance $1/N$.  The target distribution is therefore
$\mathcal{N}(\mu,1/N)$ where $\mu$ is the sample mean.  An efficient
proposal distribution has the same variance as the target distribution
$\mathcal{N}(\theta'-\theta,1/N)$, so we take $q(\theta'\mid\theta) =
\phi((\theta'-\theta)N)$, where $\phi(x)$ is the Normal density
function with zero mean and unit variance. This proposal is symmetric
$q(\theta',\theta)=q(\theta,\theta')$, and we assumed uniform
prior for $\theta$, so $\psi(u,\theta',\theta)$ reduces to $1/N\log
u$.

@@

 The choices above for $\theta$ and $\theta'$ guarantee that
  \begin{equation}
    \log u_0 \in -[0.5,1][0.75,1.5] = -[0.375,1.5]
        \end{equation}
and consider the range of $u$ values
  \begin{equation}
    \log u \in \log u_0 + [-0.5,0.375]
  \end{equation}
  the size of the range in $u$ is at least $\exp([-2,-1.125]) = [0082,0.286]$ and occurs
  with probability at least $p_2=0.2044$. With $u$ in this range, we rewrite the test as:
  \begin{equation}
    \frac{1}{b}\sum_{i=1}^b x_i-0.5 \hspace{0.1in} \not\approx \hspace{0.1in} \frac{\log u/u_0}{N(\theta'-\theta)}
  \end{equation}
  so that the LHS has expected value zero.  Given our choice of intervals for the variables, the RHS is in the range $1/\sqrt{N}[-1,0.75]$.
  The standard deviation of the LHS given the interval constraints is at least $0.5/\sqrt{b}$. And
  so the gap between LHS and RHS is at most $2\sqrt{b/N}$ standard deviations.

  The samples $\theta$, $(\theta'-\theta)$
  and $u$ are drawn independently and so the probability of the conjunction of these events is
  $c = p_0 p_1 p_2 = 0.009$.
\end{proof}

@@

For our new M-H test, we denote the exact and approximate log likelihood ratios as $\Delta$ and $\Delta^*$.
First, $\Delta$ is defined as:
\begin{equation}\label{eq:delta0}
  \Delta(\theta,\theta') = \log\frac{p(\theta')q(\theta\mid\theta')}{p(\theta)q(\theta'\mid\theta)}
\end{equation}
and in the Bayesian case
\begin{equation}\label{eq:delta1}
\Delta(x,\theta,\theta')  =
\log \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
  \theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)}
  \end{equation}
and we can separate out terms dependent and independent of the data $x$ as:
\begin{equation}\label{eq:delta2}
\Delta(x,\theta,\theta') = N\psi(1,\theta,\theta') + \sum_{i=1}^N\log\frac{p(x_i\mid\theta')}{p(x_i\mid\theta)}
\end{equation}
A minibatch estimator of $\Delta$ can be defined as:
\begin{equation}\label{eq:delta3}
    \Delta^*(x,\theta,\theta') =
N\psi(1,\theta,\theta') + \frac{N}{b}\sum_{i=1}^b\log\frac{p(x_i\mid\theta')}{p(x_i\mid\theta)}
\end{equation}
Note that $\Delta$ and $\Delta^*$ are evaluated on the full dataset and a minibatch of size $b$
respectively. The scaling term $\frac{N}{b}$ ensures that $\Delta^*(x,\theta,\theta')$ is an
unbiased estimator of $\Delta(x,\theta,\theta')$.

@@

As a sanity check, choosing $g(s) = \min\{1, \exp(s) \}$ produces the
standard Metropolis acceptance test $\alpha(\theta,\theta') = g(\Delta(\theta,\theta')) = 
\min\left\{1, \frac{p(\theta')q(\theta \mid
  \theta')}{p(\theta)q(\theta' \mid \theta)}\right\}$. $g(s)$ satisfies
the condition $g(s) = \exp(s)g(-s)$, and therefore detailed
balance. In fact it is the optimal acceptance function in terms of
acceptance rate, since it accepts with probability 1 for $\Delta > 0$.

@@

We now move to acceptance testing using the minibatch estimator
$\Delta^*(x,\theta,\theta')$. From equation (\ref{eq:delta3}),
$\Delta^*(x,\theta,\theta')$ can be represented as a constant term
plus a sum of $b$ iid terms of the form
$\log\frac{p(x_i|\theta')}{p(x_i|\theta)}$. As $N$ increases, $\Delta^*(\theta,\theta')$
therefore has a distribution which approaches a normal distribution by the central limit
theorem. Later we will provide specific bounds between the CDF of $\Delta^*(\theta,\theta')$
and a Normal CDF, but for now we use an asymptotic argument.

In the limit, since $\Delta^*$ is normally distributed about its mean $\Delta$, we
can write
\begin{equation}\label{eq:relationship}
\Delta^* = \Delta + X_{\rm norm}, \quad X_{\rm norm} \sim \mathcal{\bar{N}}(0, \sigma^2(\Delta^*)),
\end{equation}
where $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ denotes a distribution which is approximately
normal with variance $\sigma^2(\Delta^*)$. 
But to perform the test in equation (\ref{theexacttest}) we want $\Delta + X$ for
a logistic random $X$ (call it $X_{\rm log}$ from now on). 
Our approach is to decompose $X_{\rm log}$ as
\begin{equation}\label{eq:deconvolution}
X_{\rm log} = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ is a zero-mean Gaussian and $X_{\rm
  corr}$ is a zero-mean ``correction'' term with distribution $C(X)$.
The two variables are added (i.e., their distributions convolve) to
form $X_{\rm log}$. It will be convenient to express this relationship
in terms of the CDFs $N(X_{\rm norm}) = \Phi(X_{\rm norm}/\sigma)$ and
$S(X_{\rm log})$, where $\Phi(.)$ is the Normal CDF with zero mean and
unit variance and $S(.)$ is the standard logistic function.

Equation (\ref{eq:deconvolution}) implies that $S = N * C$ where $*$ denotes convolution. We will later
derive a numerical representation of $C$ by deconvolution, and using this we can generate
samples of $X_{\rm corr}$.

@@

\begin{algorithm}[t]
\Input{Number of samples $T$, minibatch size $m$, pre-computed $X_{\rm
corr}$ distribution, and initial sample $\theta_1$.}
\Output{A chain of $T$ samples $\{\theta_1, \ldots, \theta_T\}$ from $p(\theta)$\;}
\For{$t=\{1, \ldots, T\}$}{
    Propose a candidate $\theta'$\ from proposal distribution $q(\theta'\mid\theta_t)$\;
    Draw a minibatch of $m$ points $x_i$, compute $\Delta^*(x,\theta_t,\theta')$ and sample variance $\sigma^2(\Delta^*)$\;
    \While {$\sigma^2(\Delta^*) \ge 1$} {
    	Draw $m$ more samples to augment the current minibatch, and update $\Delta^*$ and $\sigma^2(\Delta^*)$ \;
    }
    Let $\sigma_n^2 = 1 - \sigma^2(\Delta^*)$, draw a normal variable $X_n \in \mathcal{N}(0,\sigma_n^2)$ and
    $X_c$ from the correction distribution\;

        \eIf{$\Delta^* + X_n + X_c > 0$}{
            Accept the candidate, $\theta_{t+1} = \theta'$\;
        }{
            Reject and re-use the old sample, $\theta_{t+1} = \theta_t$\;
        }

@@

The outline of this section is as follows: we first present bounds on the absolute
and relative error (in CDF) of the distribution of $\Delta^*$ vs a Normal distribution. We then show that
these errors bounds are preserved after the addition of other random variables, in particular
$X_n$ and $X_c$. From this it follows that the acceptance test has the same error bound.

@@

To apply this to our test, let $\mu = \frac{1}{N}\sum_{i=1}^N\log\frac{p(x_i|\theta')}{p(x_i|\theta)}$ which is
the expected log data probability ratio across the entire dataset. Then
Let $X_i = N(\log\frac{p(x_i|\theta')}{p(x_i|\theta)}-\mu)$, are zero-mean, IID variables.
Take $n=b$ the minibatch size, and then:
\begin{equation}
  \Delta^*(x,\theta,\theta') = \hat{X} + \Delta(x,\theta,\theta')
\end{equation}
so that $\sigma(\hat{X}) = \sigma(\Delta^*)$.
\begin{corollary}
  We can now substitute into equation (\ref{eq:clt-bounds2}) and displace by the mean, giving:
\begin{equation}\label{eq:clt-bounds3}
   \sup_y\left|{\rm Pr}(\Delta^*< y) - \Phi\left(\frac{y-\Delta}{\sigma(\Delta^*)}\right)\right| \leq \frac{6.4E|X|^3+2E|X|}{\sqrt{b}} = \epsilon(\theta,\theta',b)
\end{equation}
\end{corollary}

We have shown that the distribution of $\Delta^*$ approximates a
Normal distribution with mean $\Delta$ and variance
$\sigma^2(\Delta^*)$, and bounded its error with estimable
quantities. Both $E|X|$ and $E|X|^3$ can be estimated as means of
$|X_i - \hat{X}|$ and $|X_i - \hat{X}|^3$ respectively on each minibatch. We expect this
will often be accurate enough on minibatches with hundreds or
thousands of points, but otherwise bootstrap CIs can be computed from
those sequences. Since the bounds are monotone in $E|X|$ and $E|X|^3$,
using upper bootstrap CI limits will provide high-confidence error bounds.

>>>>

In the Metropolis-Hastings method~\cite{gilks1996markov,brooks2011handbook}, a difficult-to-compute
probability distribution $p(\theta)$ is sampled using a Markov chain $\theta_1,\ldots,\theta_n$. The 
sample $\theta_{t+1}$ at time $t+1$ is generated using a candidate $\theta'$ from a (simpler) proposal
distribution $q(\theta'\mid \theta_t)$, filtered by an acceptance test. The acceptance test is usually
a Metropolis test. The Metropolis test has acceptance probability:
\begin{equation}\label{eq:traditional}
\alpha(\theta_t,\theta') = \frac{p(\theta')q(\theta_t \mid \theta')}{p(\theta_t)q(\theta' \mid \theta_t)} \wedge 1
\end{equation}
where $a \wedge b$ denotes $\min(a,b)$.  With probability $\alpha(\theta_t,\theta')$, we accept $\theta'$
and set $\theta_{t+1} = \theta'$, otherwise set
$\theta_{t+1}=\theta_t$.  The test is often implemented with an
auxiliary random variable $u \sim \mathcal{U}(0,1)$ with a comparison $u < \alpha(\theta_t,\theta')$. $\mathcal{U}(a,b)$ denotes the uniform distribution on the interval $[a,b]$. 
For notational simplicity, from now on we will drop the subscript $t$ for the current sample $\theta_t$ and
denote is simply as $\theta$.

The acceptance test guarantees detailed balance, which means
\begin{equation}\label{detailed_balance1}
  p(\theta)p(\theta'\mid\theta) = p(\theta')p(\theta \mid\theta')
\end{equation}
where $p(\theta'\mid\theta)$ is the probability of a transition from state $\theta$ to $\theta'$. Here
$p(\theta'\mid\theta) = q(\theta'\mid\theta)\alpha(\theta,\theta')$ so the detailed balance equation becomes
\begin{equation}\label{detailed_balance2}
  p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta)
\end{equation}
The detailed balance condition, together with ergodicity, guarantee that the
Markov chain has a unique stationary distribution $\pi(\theta) = p(\theta)$.

For Bayesian inference, we would like to sample from a parameter distribution for $\theta$ based on some
observed data $x_1,\ldots,x_N$. i.e. we want to sample from 
$p(\theta \mid x_1, \ldots, x_N)$. The acceptance probability is now:
\begin{equation}
\alpha(\theta,\theta') = 
 \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
\theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)} \wedge 1
\end{equation}
where $p_0(\theta)$ is a prior, and $p(x_i \mid \theta)$ are the
probabilities of the observations. Computing samples this way requires
the use of all $N$ training data points, but this is very expensive
for large datasets. To address this challenge,~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate Metropolis-Hasting tests using sequential hypothesis
testing. During each iteration, they start with a small minibatch of
data and test the hypothesis that the sample $\theta'$ based on an
approximate version of the test $u < \alpha(\theta,\theta')$. If the
approximate test cannot make a decision with sufficient confidence,
then the minibatch size is increased and the test repeats. This
process continues until a decision. The bounds depend on either a
asymptotic central limit theorem \cite{cutting_mh_2014} or a
concentration bound \cite{icml2014c1_bardenet14}, the latter requiring
direct bounds on the log likelihood ratio.  The problem with both
methods is the expense of resolving small differences.  In the worst
case, all $N$ data points may be needed, and as we show next this
worst case can occur about $\Omega(N)$ times during the performance of
$N$ tests.

Following ~\cite{icml2014c1_bardenet14}, we write the test $u < \alpha(\theta,\theta')$ in
the equivalent form $\Lambda(\theta,\theta') > \psi(u,\theta,\theta')$. Our definitions differ
from those in \cite{icml2014c1_bardenet14} by a factor of $N$ to simplify our analysis later on. 
\begin{equation}
\Lambda(\theta,\theta') = \sum_{i=1}^N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\hspace{0.2in} {\rm and} \hspace{0.2in}
\psi(u,\theta,\theta') = \log\left(u\frac{q(\theta'|\theta)p_0(\theta)}{q(\theta|\theta')p_0(\theta')}\right)
\end{equation}

@@

\begin{equation}
\Lambda^*(\theta,\theta') = \frac{N}{b}\sum_{i=1}^b \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}

Finally, it will be convenient in what follows to define the individual components that
contribute to the sums above:

\begin{equation}\label{eq:individual_terms}
\Lambda_i(\theta,\theta') = N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}

So $\Lambda(\theta,\theta')$ is the mean of $\Lambda_i(\theta,\theta')$ over the
entire dataset, and $\Lambda^*(\theta,\theta')$ is the mean over the minibatch
of $\Lambda_i(\theta,\theta')$.

Since the minibatch contains randomly selected samples $x_i$, the
values $\Lambda_i$ are independent, IID random variables.  By the central
limit theorem, we expect $\Lambda^*(\theta,\theta')$ to be
approximately Normal.  The acceptance test then becomes a statistical
test of the hypothesis that $\Lambda(\theta,\theta') >
\psi(u,\theta,\theta')$ by establishing that
$\Lambda^*(\theta,\theta')$ is substantially larger than
$\psi(u,\theta,\theta')$.  In ~\cite{cutting_mh_2014} an asymptotic
central limit argument was used to derive this gap, while in
\cite{icml2014c1_bardenet14} a concentration bound was used. In both
cases, the resulting tests were shown to give useful reductions in
number of samples required over using the full dataset. But no
worst-case bounds were given on average number of samples needed.

@@

\subsection{An Example}
Consider a simple model where samples $x_i$ for $i=1,\ldots,N$ are
drawn from a Normal distribution $\mathcal{N}(\theta,1)$ where
$\theta=0.5$, and with a uniform prior on $\theta$. The log likelihood ratio is
\begin{equation}
\Lambda^*(\theta,\theta') =  \frac{N}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta)}=
  %%\frac{1}{2}\left( (x_i-\theta')^2 - (x_i - \theta)^2\right) =
  N(\theta'-\theta)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.
Since the $x_i$ have unit variance, their mean has variance $1/b$, and the variance $\sigma^2(\Lambda^*)$
of $\Lambda^*(\theta,\theta')$ is $(\theta'-\theta)^2N/b$. In order to pass
a hypothesis test that $\Lambda > \psi$, we will need to establish a
large enough gap (several $\sigma(\Lambda^*)$) between $\Lambda^*(\theta',\theta)$
and $\psi(u,\theta',\theta)$.

We would like to sample from the posterior distribution of $\theta$
which is a normal distribution centered on the sample mean, and with
variance $1/N$.  The target distribution is therefore
$\mathcal{N}(\mu,1/N)$ where $\mu$ is the sample mean, and $\mathcal{N}(\mu,\sigma^2)$ is a
normal distribution with mean $\mu$ and variance $\sigma^2$.  An efficient
proposal distribution has the same variance as the target distribution
$\mathcal{N}(\theta'-\theta,1/N)$, so we take $q(\theta'\mid\theta) =
\phi((\theta'-\theta)N)$, where $\phi(x)$ is the Normal density
function with zero mean and unit variance. This proposal is symmetric
$q(\theta'\mid\theta)=q(\theta\mid\theta')$, and we assumed uniform
prior for $\theta$, so $\psi(u,\theta',\theta)$ reduces to $\log
u$.

@@

 The choices above for $\theta$ and $\theta'$ guarantee that
  \begin{equation}
    \log u_0 \in -[0.5,1][0.75,1.5] = -[0.375,1.5]
        \end{equation}
and consider the range of $u$ values
  \begin{equation}
    \log u \in \log u_0 + [-0.5,0.375]
  \end{equation}
  the size of the range in $u$ is at least $\exp([-2,-1.125]) = [0.13534,0.32465]$ and occurs
  with probability at least $p_2=0.18932$. With $u$ in this range, we rewrite the test as:
  \begin{equation}
    \frac{1}{b}\sum_{i=1}^b x_i-0.5 \hspace{0.1in} \not\approx \hspace{0.1in} \frac{\log u/u_0}{N(\theta'-\theta)}
  \end{equation}
  so that the LHS has expected value zero.  Given our choice of intervals for the variables, the RHS is in the range $1/\sqrt{N}[-1,0.75]$.
  The standard deviation of the LHS given the interval constraints is at least $0.5/\sqrt{b}$. And
  so the gap between LHS and RHS is at most $2\sqrt{b/N}$ standard deviations.

  The samples $\theta$, $(\theta'-\theta)$
  and $u$ are drawn independently and so the probability of the conjunction of these events is
  $c = p_0 p_1 p_2 = 0.0085$.
\end{proof}

@@

For our new M-H test, we denote the exact and approximate log likelihood ratios as $\Delta$ and $\Delta^*$.
First, $\Delta$ is defined as:
\begin{equation}\label{eq:delta0}
  \Delta(\theta,\theta') = \log\frac{p(\theta')q(\theta\mid\theta')}{p(\theta)q(\theta'\mid\theta)}
\end{equation}
and in the Bayesian case
\begin{equation}\label{eq:delta1}
\Delta(\theta,\theta')  =
\log \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
  \theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)}
  \end{equation}
and we can separate out terms dependent and independent of the data $x$ as:
\begin{equation}\label{eq:delta2}
  \Delta(\theta,\theta') = \sum_{i=1}^N\log\frac{p(x_i\mid\theta')}{p(x_i\mid\theta)} - \psi(1,\theta,\theta') =
  \Lambda(\theta,\theta') -  \psi(1,\theta,\theta')
\end{equation}
A minibatch estimator of $\Delta$ can be defined as:
\begin{equation}\label{eq:delta3}
    \Delta^*(\theta,\theta') =
\frac{N}{b}\sum_{i=1}^b\log\frac{p(x_i\mid\theta')}{p(x_i\mid\theta)} - \psi(1,\theta,\theta') =
\Lambda^*(\theta,\theta') -     \psi(1,\theta,\theta')
\end{equation}
Note that $\Delta$ and $\Delta^*$ are evaluated on the full dataset and a minibatch of size $b$
respectively. The scaling term $\frac{N}{b}$ ensures that $\Delta^*(\theta,\theta')$ is an
unbiased estimator of $\Delta(\theta,\theta')$.

@@

As a sanity check, choosing $g(s) = \exp(s) \wedge 1$ produces the
standard Metropolis acceptance test $\alpha(\theta,\theta') = g(\Delta(\theta,\theta')) = 
\frac{p(\theta')q(\theta \mid
  \theta')}{p(\theta)q(\theta' \mid \theta)}\wedge 1$. $g(s)$ satisfies
the condition $g(s) = \exp(s)g(-s)$, and therefore detailed
balance. In fact it is the optimal acceptance function in terms of
acceptance rate, since it accepts with probability 1 for $\Delta > 0$.

@@

We now move to acceptance testing using the minibatch estimator
$\Delta^*(\theta,\theta')$. From equation (\ref{eq:delta3}),
$\Delta^*(\theta,\theta')$ can be represented as a constant term
plus the mean of $b$ IID terms $\Lambda_i(\theta,\theta')$ of the form
$N\log\frac{p(x_i|\theta')}{p(x_i|\theta)}$. As $b$ increases, $\Delta^*(\theta,\theta')$
therefore has a distribution which approaches a normal distribution by the central limit
theorem. Later we will provide specific bounds between the CDF of $\Delta^*(\theta,\theta')$
and a Normal CDF, but for now we use an asymptotic argument.

In the limit, since $\Delta^*$ is normally distributed about its mean $\Delta$, we
can write
\begin{equation}\label{eq:relationship}
\Delta^* = \Delta + X_{\rm norm}, \quad X_{\rm norm} \sim \mathcal{\bar{N}}(0, \sigma^2(\Delta^*)),
\end{equation}
where $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ denotes a distribution which is approximately
normal with variance $\sigma^2(\Delta^*)$. 
But to perform the test in equation (\ref{theexacttest}) we want $\Delta + X$ for
a logistic random $X$ (call it $X_{\rm log}$ from now on). 
Our approach is to decompose $X_{\rm log}$ as
\begin{equation}\label{eq:deconvolution}
X_{\rm log} = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ has zero-mean, Normal distribution with
variance $\sigma^2$, and $X_{\rm corr}$ is a zero-mean ``correction''
term with density $C_{\sigma}(X)$.  The two variables are added
(i.e., their distributions convolve) to form $X_{\rm log}$. It will be
convenient to express this relationship in terms of the CDFs $N_{\sigma}(X_{\rm
  norm}) = \Phi(X_{\rm norm}/\sigma)$ and $S(X_{\rm log})$, where
$\Phi(.)$ is the Normal CDF with zero mean and variance $\sigma^2$ and
$S(.)$ is the standard logistic function.

Equation (\ref{eq:deconvolution}) implies that $S = N_{\sigma} * C_{\sigma}$ where $*$ denotes convolution. We will later
derive a numerical representation of $C$ by deconvolution, and using this we can generate
samples of $X_{\rm corr}$.

@@

\begin{algorithm}[t]
\Input{Number of samples $T$, minibatch size $m$, error bound $\delta$, pre-computed correction $C_1(X)$ distribution, initial sample $\theta_1$.}
\Output{A chain of $T$ samples $\{\theta_1, \ldots, \theta_T\}$ from $p(\theta)$\;}
\For{$t=\{1, \ldots, T\}$}{
    Propose a candidate $\theta'$\ from proposal distribution $q(\theta'\mid\theta_t)$\;
    Draw a minibatch of $m$ points $x_i$, compute $\Delta^*(\theta_t,\theta')$ and sample variance $\sigma^2(\Delta^*)$\;
    Estimate moments $E|\Lambda_i-\Lambda|$ and $E|\Lambda_i-\Lambda|^3$ from the sample, and error $\epsilon$ from equation (\ref{eq:clt-bounds3})\;
    \While {$\sigma^2(\Delta^*) \ge 1$ {\bf or} $\epsilon > \delta$}{
    	Draw $m$ more samples to augment the minibatch, update $\Delta^*$, $\sigma^2(\Delta^*)$ and $\epsilon$ estimates\;
    }
    Draw $X_{\rm norm} \in \mathcal{N}(0,1-\sigma^2(\Delta*))$ and
    $X_{\rm corr}$ from the correction distribution $C_1(X)$\;

        \eIf{$\Delta^* + X_{\rm norm} + X_{\rm corr} > 0$}{
            Accept the candidate, $\theta_{t+1} = \theta'$\;
        }{
            Reject and re-use the old sample, $\theta_{t+1} = \theta_t$\;
        }

@@

The outline of this section is as follows: we first present bounds on the absolute
and relative error (in CDF) of the distribution of $\Delta^*$ vs a Normal distribution. We then show that
these errors bounds are preserved after the addition of other random variables, in particular
$X_{\rm norm}$ and $X_{\rm corr}$. From this it follows that the acceptance test has the same error bound.

@@

To apply this to our test, let $X_i = \Lambda_i(\theta,\theta') - \Lambda(\theta,\theta')$,
then the $X_i$ are zero-mean, IID variables.
Take $n=b$ the minibatch size, and then:
\begin{equation}
\hat{X} = \Delta^*(\theta,\theta') - \Delta(\theta,\theta') 
\end{equation}
so that $\sigma(\hat{X}) = \sigma(\Delta^*)$.
\begin{corollary}
  We can now substitute into equation (\ref{eq:clt-bounds2}) and displace by the mean, giving:
\begin{equation}\label{eq:clt-bounds3}
   \sup_y\left|{\rm Pr}(\Delta^*< y) - \Phi\left(\frac{y-\Delta}{\sigma(\Delta^*)}\right)\right| \leq \frac{6.4E|X|^3+2E|X|}{\sqrt{b}} = \epsilon(\theta,\theta',b)
\end{equation}
\end{corollary}

We have shown that the distribution of $\Delta^*$ approximates a
Normal distribution with mean $\Delta$ and variance
$\sigma^2(\Delta^*)$, and bounded its error with estimable
quantities. Both $E|X|$ and $E|X|^3$ can be estimated as means of
$|\Lambda_i - \Lambda|$ and $|\Lambda_i - \Lambda|^3$ respectively on each minibatch. We expect this
will often be accurate enough on minibatches with hundreds or
thousands of points, but otherwise bootstrap CIs can be computed from
those sequences. Since the bounds are monotone in $E|X|$ and $E|X|^3$,
using upper bootstrap CI limits will provide high-confidence error bounds.
